Mechanising Recursion Schemes with Magic-Free Coq Extraction

### 2024-03-18 paper/paper.tex
Introduction fits in the page

However, when used in the context of languages with general recursion like
(e.g.) Haskell, recursion schemes cannot ensure termination, but while formalising
recursion schemes in a type theory does provide stronger termination guarantees, to the best of our knowledge, not many attempts have been made at
mechanise structured recursion schemes in a proof assistant.

Recently, Abreu et al.~\cite{AbreuDHJMS23} encoded an algebraic approach to divide-and-conquer
computations in which termination is entirely enforced by the typing discipline. Their approach solves the problem of
termination proofs as well as the performance of the code that is run
\emph{within Coq}, but it does not allow the extraction of idiomatic OCaml code.
By this we refer to code that (1) does not preserve the recursive structure of
common implementations; and (2) lead to unsafe casts like \ocaml{Obj.magic} in the
generated code.  In OCaml, \ocaml{Obj.magic} is problematic in that it has been proven that,
for higher-order programs, simple interoperations can lead to incorrect
behaviour or even segfaults~\cite{forster:hal-04329663} and, more generally, it invalidates the fast-and-loose principle~\cite{DanielssonHJG06}. % The absence of \ocaml{Obj.magic} provides even stronger
% guarantees to the extracted code. In particular, the extracted code from Coq is
% well-defined allowing the programmer to use the so-called \emph{fast-and-loose
% reasoning}~\cite{DanielssonHJG06} on the extracted code. % For example, the
% % programmer would need to manually
% %
% check that the inputs satisfy the necessary assumptions which would be
% % impossible to enforce in the presence of unsafe casts. Overall, the absence
% % of \ocaml{Obj.magic} makes it easer to reason about, and interoperate with
% % extracted code.

This work presents the first formalisation in Coq of \emph{hylomorphisms}, as
well as their algebraic laws that: (1) is \emph{axiom-free}; and (2) allows the
extraction of idiomatic OCaml code. While programmers still need to reason about
the termination of their programs, our mechanisation will allow the use of the
algebraic laws of hylomorphisms for program reasoning, as well as extracting to
idiomatic code. As an example our mechanisation of \ocaml{qsort} as a
hylomorphism will produce the following code:
\begin{coqcode}
let rec qsort = function | [] -> []
  | h :: t -> let (l, r) = partition (fun x0 -> (<=) x0 h) t in
              let cx = fun p -> match p with | Lbranch -> l | Rbranch -> r in 
              app (qsort (cx Lbranch)) (h :: (qsort (cx Rbranch)))
\end{coqcode}
The contributions we make in this paper are as follows:
\begin{itemize}
  \item We provide a framework for \emph{generic programming} with recursion
    schemes in Coq, with a proof of their algebraic laws and 
    program equivalences, that allows clean code extraction.
  \item We formalise divide and conquer, dynamic programming, and
    mutual recursion algorithms.
  \item We verified the short-cut fusion optimisation in Coq and
    extract optimised code to OCaml.
\end{itemize}

@@

\section{Mechanising Extractable Container Functors}
In order to abstract recursion patters we need to be able to abstract away from
the particular shape of the data.  This is achieved by introducing the concept
of \emph{functors} which have the suitable fixed-point properties, i.e.\ those
who have a initial algebras. A common approach to construct such functors is to
use \emph{containers}~\cite{AbbottAG05} (Section~\ref{sec:containers}).
However, reasoning about container equality will require us to consider both
functional extensionality and heterogeneous equality. We avoid these axioms by
introducing a custom equivalence relation on types
(Section~\ref{sec:types-morphisms}).

\subsection{Functors and Containers}
\label{sec:containers}
Functors are functions \coq{F : Type -> Type} which additionally have a
map
\begin{coqcode}
fmap : forall A B, (A -> B) -> F A -> F B
\end{coqcode}
witnessing the idea that a
functor represents a container for abstract data that can be manipulated without
changing the outer structure. For example, the type \coq{List : Type -> Type} is
a functor and \coq{List A} is the type of lists over elements of type \coq{A}
with the obvious \coq{fmap} function recursively traversing a list and applying
the function \coq{A -> B} for each element. Additionally, the type of
lists \coq{List A} arises as the least fixed-point of the
functor \coq{F X = unit + (A * X) }.

@@

\section{Formalising Recursion Schemes}
Recursion schemes provide an abstract way to consume and generate data. We now
proceed onto describing how to formalise hylomorphisms in Coq. We first
formalise algebras for container extensions (Section~\ref{sec:algebras}), then
we formalise coalgebras (Section~\ref{sec:coalg}) and then we put together
these notions to formalise recursive coalgebras and hylomorphisms
(Section~\ref{sec:rec-coalgebras}).

@@

\section{Extraction}
We go in this section through a series of case studies of various recursive
algorithms.  We show how they can be encoded in our framework, how can we do
program calculation techniques for optimisation, and how can they be extracted
to idiomatic OCaml code. Our examples are the Quicksort and Mergesort
algorithms (Section~\ref{sec:sorting}), dynammic programming and Knapsack
(Section~\ref{sec:knap}), and examples of the shortcut deforestation
optimisation in our framework (Section~\ref{sec:shortcut}).
%The steps are the following:
%\begin{enumerate}
%  \item Write a specification as a morphism \coq{f : A ~> B}.
%  \item Use Coq's generalised rewriting and the laws of morphisms to 
%    rewrite it to some \coq{g : A ~> B} such that \coq{f =e g}.
%  \item Project the function \coq{h : A -> B} from \coq{g}
%    (i.e. \coq{h = app g}).
%  \item Extract \coq{h} to OCaml.
%\end{enumerate}
%Note that steps 2 and 3 are \emph{proofs} that \coq{f} and \coq{h} are
%\emph{extensionally equal}. Then, Coq's extraction mechanism will erase any
%proofs from the code, and inline our definitions. To ease the application of
%these steps, we provide the following helper type:
%\begin{coqcode}
%  Record Ext (f : A ~> B) := 
%    MkExt { target :> A -> B; tgt_eq : app f =e target; }.
%\end{coqcode}
%Note that \coq{tgt_eq} will be erased when extracting \coq{Ext f} to OCaml,
%since it is in \coq{Prop}. This implies that the code extracted for \coq{Ext f}
%will be the function \coq{target} that is extensionally equal to \coq{f}.  We
%use a shortcut tactic, \coq{calculate}, that is a synonym to 
%\coq{eapply MkExt}. As a result of this tactic, we will be given the goal 
%\coq{f =e ?Goal0}. We can rewrite \coq{f}, unfold definitions, etc, and
%complete the definition with \coq{reflexivity}, and this will instantiate the
%metavariable \coq{?Goal0} to our desired function. We will use this extensively
%in our examples.
%\mvol{This is a very dry overview that doesn't need to be here. We can just jump
%  right into the examples.}
%
%\mvol{Instead of explaining what types, tactics, etc., will be used, this space should
%  be used for a little overview of the rest of the section, like: ``We will go over a series
%  of case studies of various recursive algorithms, showing how they can be encoded
%  in our framework and extracted to idiomatic OCaml code.''}
%
%\mvol{There are three sub-sections: divide-and-conquer, dynamic programming, and deforestation.
%  Each of these is written like a case study, with a particular program worked out all the way
%  to extracted OCaml code. The sub-section titles should reflect this, and there needs to be
%  a sentence at the beginning of each section re-iterating this to make the whole thing
%  flow more smoothly.}

@@

\section{Related Work}
%
%\begin{itemize}
%  \item McBride devised a generalisation of Bove and Capretta's general
%recursive function monad~\cite{McBride15, BoveC01}
%  \item Guarded recursion for productive definitions and general recursion
%~\cite{AtkeyM13, PaviottiMB15}
%  \item PACO library~\cite{HurNDV13}
%  \item Sized Types \cite{HughesPS96, abs-1202-3496}
%  \item Copatterns \cite{AbelPTS13, AbelP16}
%  \item Not that related: extraction for mutable arrays~\cite{SAKAGUCHI2020102372}
%        \item Extraction of Distributed Mobile code in Coq~\cite{MiculanP12}.
%\end{itemize}
Encoding recursion schemes in Coq is not new. We compare our work with other
encodings of program calculation techniques in Coq (Section~\ref{sec:calccoq}),
recursion schemes in Coq (Sections~\ref{sec:calccoq} and~\ref{sec:reccoq}).
Finally, we compare our encoding of recursion schemes to other forms of
termination checking (Section~\ref{sec:termcheck}, and sized types
(Section~\ref{sec:sizedT}), as a way to guarantee termination of nonstructural
recursion.

>>>>

However, when used in the context of languages with general recursion like
(e.g.) Haskell, recursion schemes cannot ensure termination, but while
formalising recursion schemes in a type theory does provide stronger termination
guarantees, to the best of our knowledge, not many attempts have been made at
mechanising structured recursion schemes.

Recently, Abreu et al.~\cite{AbreuDHJMS23} encoded an algebraic approach to
divide-and-conquer computations in which termination is entirely enforced by the
typing discipline. Their approach solves the problem of termination proofs as
well as the performance of the code that is run \emph{within Coq}, but it does
not allow the extraction of idiomatic OCaml code.  By this we refer to code that
(1) does not preserve the recursive structure of common implementations; and (2)
lead to unsafe casts like \ocaml{Obj.magic} in the generated code.  This latter is problematic in that, for higher-order programs, simple interoperations can lead to incorrect behaviour or
even segfaults~\cite{forster:hal-04329663} and, moreover, it invalidates
the fast-and-loose principle~\cite{DanielssonHJG06}. % The absence
% of \ocaml{Obj.magic} provides even stronger
% guarantees to the extracted code. In particular, the extracted code from Coq is
% well-defined allowing the programmer to use the so-called \emph{fast-and-loose
% reasoning}~\cite{DanielssonHJG06} on the extracted code. % For example, the
% % programmer would need to manually
% %
% check that the inputs satisfy the necessary assumptions which would be
% % impossible to enforce in the presence of unsafe casts. Overall, the absence
% % of \ocaml{Obj.magic} makes it easer to reason about, and interoperate with
% % extracted code.

This work presents the first formalisation in Coq of \emph{hylomorphisms}, as
well as their algebraic laws that: (1) is \emph{axiom-free}; and (2) allows the
extraction of idiomatic OCaml code. While programmers still need to reason about
the termination of their programs, our mechanisation will allow the use of the
algebraic laws of hylomorphisms for program reasoning, as well as extracting to
idiomatic code. As an example our mechanisation of \ocaml{qsort} as a
hylomorphism will produce the following code:
\begin{coqcode}
let rec qsort = function | [] -> []
  | h :: t -> let (l, r) = partition (fun x0 -> (<=) x0 h) t in
              let cx = fun p -> match p with | Lbranch -> l | Rbranch -> r in 
              app (qsort (cx Lbranch)) (h :: (qsort (cx Rbranch)))
\end{coqcode}

Finally, in this paper we provide a framework for \emph{generic programming} with recursion
schemes in Coq, with a proof of their algebraic laws and program equivalences for clean code extraction:
\begin{itemize}
  \item In Section~\ref{sec:containers} we formalise the type of container functors suitably modified for program extraction
  \item In Section~\ref{sec:recursion-schemes} we formalise hylomorhisms and prove their universal property
  \item In Section~\ref{sec:examples} we provide examples of divide and conquer, dynamic programming, and mutual recursion algorithms and we verify the short-cut fusion optimisation and show the extracted optimised code to OCaml.
\end{itemize}

@@

\section{Mechanising Extractable Container Functors}
\label{sec:containers}
In order to abstract recursion patters we need to be able to abstract away from
the particular shape of the data.  This is achieved by introducing the concept
of \emph{functors} which have the suitable fixed-point properties, i.e.\ those
who have a initial algebras. A common approach to construct such functors is to
use \emph{containers}~\cite{AbbottAG05} (Section~\ref{sec:functors}).
However, reasoning about container equality will require us to consider both
functional extensionality and heterogeneous equality. We avoid these axioms by
introducing a custom equivalence relation on types
(Section~\ref{sec:types-morphisms}).

\subsection{Functors and Containers}
\label{sec:functors}
Functors are functions \coq{F : Type -> Type} which additionally have a
map
\begin{coqcode}
fmap : forall A B, (A -> B) -> F A -> F B
\end{coqcode}
witnessing the idea that a
functor represents a container for abstract data that can be manipulated without
changing the outer structure. For example, the type \coq{List : Type -> Type} is
a functor and \coq{List A} is the type of lists over elements of type \coq{A}
with the obvious \coq{fmap} function recursively traversing a list and applying
the function \coq{A -> B} for each element. Additionally, the type of
lists \coq{List A} arises as the least fixed-point of the
functor \coq{F X = unit + (A * X) }.

@@

\section{Formalising Recursion Schemes}
\label{sec:recursion-schemes}
Recursion schemes provide an abstract way to consume and generate data. We now
proceed onto describing how to formalise hylomorphisms in Coq. We first
formalise algebras for container extensions (Section~\ref{sec:algebras}), then
we formalise coalgebras (Section~\ref{sec:coalg}) and then we put together
these notions to formalise recursive coalgebras and hylomorphisms
(Section~\ref{sec:rec-coalgebras}).

@@

\section{Extraction}
\label{sec:examples}
We go in this section through a series of case studies of various recursive
algorithms.  We show how they can be encoded in our framework, how can we do
program calculation techniques for optimisation, and how can they be extracted
to idiomatic OCaml code. Our examples are the Quicksort and Mergesort
algorithms (Section~\ref{sec:sorting}), dynammic programming and Knapsack
(Section~\ref{sec:knap}), and examples of the shortcut deforestation
optimisation in our framework (Section~\ref{sec:shortcut}).
%The steps are the following:
%\begin{enumerate}
%  \item Write a specification as a morphism \coq{f : A ~> B}.
%  \item Use Coq's generalised rewriting and the laws of morphisms to 
%    rewrite it to some \coq{g : A ~> B} such that \coq{f =e g}.
%  \item Project the function \coq{h : A -> B} from \coq{g}
%    (i.e. \coq{h = app g}).
%  \item Extract \coq{h} to OCaml.
%\end{enumerate}
%Note that steps 2 and 3 are \emph{proofs} that \coq{f} and \coq{h} are
%\emph{extensionally equal}. Then, Coq's extraction mechanism will erase any
%proofs from the code, and inline our definitions. To ease the application of
%these steps, we provide the following helper type:
%\begin{coqcode}
%  Record Ext (f : A ~> B) := 
%    MkExt { target :> A -> B; tgt_eq : app f =e target; }.
%\end{coqcode}
%Note that \coq{tgt_eq} will be erased when extracting \coq{Ext f} to OCaml,
%since it is in \coq{Prop}. This implies that the code extracted for \coq{Ext f}
%will be the function \coq{target} that is extensionally equal to \coq{f}.  We
%use a shortcut tactic, \coq{calculate}, that is a synonym to 
%\coq{eapply MkExt}. As a result of this tactic, we will be given the goal 
%\coq{f =e ?Goal0}. We can rewrite \coq{f}, unfold definitions, etc, and
%complete the definition with \coq{reflexivity}, and this will instantiate the
%metavariable \coq{?Goal0} to our desired function. We will use this extensively
%in our examples.
%\mvol{This is a very dry overview that doesn't need to be here. We can just jump
%  right into the examples.}
%
%\mvol{Instead of explaining what types, tactics, etc., will be used, this space should
%  be used for a little overview of the rest of the section, like: ``We will go over a series
%  of case studies of various recursive algorithms, showing how they can be encoded
%  in our framework and extracted to idiomatic OCaml code.''}
%
%\mvol{There are three sub-sections: divide-and-conquer, dynamic programming, and deforestation.
%  Each of these is written like a case study, with a particular program worked out all the way
%  to extracted OCaml code. The sub-section titles should reflect this, and there needs to be
%  a sentence at the beginning of each section re-iterating this to make the whole thing
%  flow more smoothly.}

@@

\section{Related Work}
\label{sec:related-work}
%
%\begin{itemize}
%  \item McBride devised a generalisation of Bove and Capretta's general
%recursive function monad~\cite{McBride15, BoveC01}
%  \item Guarded recursion for productive definitions and general recursion
%~\cite{AtkeyM13, PaviottiMB15}
%  \item PACO library~\cite{HurNDV13}
%  \item Sized Types \cite{HughesPS96, abs-1202-3496}
%  \item Copatterns \cite{AbelPTS13, AbelP16}
%  \item Not that related: extraction for mutable arrays~\cite{SAKAGUCHI2020102372}
%        \item Extraction of Distributed Mobile code in Coq~\cite{MiculanP12}.
%\end{itemize}
Encoding recursion schemes in Coq is not new. We compare our work with other
encodings of program calculation techniques in Coq (Section~\ref{sec:calccoq}),
recursion schemes in Coq (Sections~\ref{sec:calccoq} and~\ref{sec:reccoq}).
Finally, we compare our encoding of recursion schemes to other forms of
termination checking (Section~\ref{sec:termcheck}, and sized types
(Section~\ref{sec:sizedT}), as a way to guarantee termination of nonstructural
recursion.
